---
title: "Class 11: Introduction to Genome Informatics Lab"
author: "Audrey Nguyen"
format: gfm
---

# Section 1. Proportion og G/G in a population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39853598-39936599;v=rs8067378;vdb=variation;vf=105535077#373531_tablePanel

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (1).csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood astham is more frequent in the GBR population than the MKL population. 

Let's now dig into this further. 

## Population Scale Analysis

> Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

Let's make a boxplot.

```{r}
ggplot(expr) + aes(geno, exp, fill = geno) + geom_boxplot(notch = TRUE)
```

It looks like the SNP between the G and A affects expression levels. 
A/A median expression ~ 32
A/G median expression = 25
G/G median expression = 20
A/A has more expression than A/G and G/G, so it looks like the A allele is induces more expression. 
